---
title: "Accessing annotations in Bioconductor"
author: "[Michael Love](http://mikelove.github.io)"
output: html_document
---

Here we will show how to use *AnnotationHub* and annotation packages
in Bioconductor to access both experimental data, and information
about things like genes. *AnnotationHub* is a *relatively* new method
for accessing experimental data and annotation information (introduced
in 2013). Previously, the main ways to access annotation information
were to download a large package with all the information contained,
or to query a web resource called *BioMart* using the *biomaRt*
Bioconductor package. *AnnotationHub* makes it much easier to look up
and download annotation information within R.

One important usage note: *AnnotationHub* creates a hidden folder in
your home directory, `~/.AnnotationHub` which accumulates the data
objects you download with the package. In the next topic, we will show
how to download the human genome which is ~700 Mb, and so you should
know that it is sitting in this hidden folder, in case you need to
clear up space later.

```{r results="hide"}
library("AnnotationHub")
ah <- AnnotationHub()
```

One way to see what kind of information is available is to put up an
HTML browser using the `display` function. 
Search for **species:** "Homo sapiens" and **description:**
"chip-seq". (I've noticed that using the global search is much slower
than using the individual fields.) When you are finished, and want to
return to R, you should hit `Esc` (or `Ctrl-C Ctrl-C` in an R terminal).

```{r eval=FALSE}
display(ah)
```

We can look up specific datasets of interest, either by selecting them
in the web browser and telling the browser to return these rows to R,
or by using the single square bracket with the `AH` code. This just
tells us about the record:

```{r}
ah["AH28856"]
```

If we want to actually download this dataset (a set of ChIP-seq peaks
for the histone modification H3K4me1), we use the double square
brackets. The data are the set of locations in the genome where the
histone modification was detected. Each range is associated with a
score, a "signal value", a p-value and a q-value.


```{r}
peaks <- ah[["AH28856"]]
peaks
```

We have the chromosome and genome information as well:

```{r}
seqinfo(peaks)
```

# Organism databases

Another type of annotation we can pull down using *AnnotationHub* is
an *OrgDb* or organism database. We subset the hub to only those
objects that are of type *OrgDb*, and then we query for Homo sapiens.


```{r}
orgs <- subset(ah, ah$rdataclass == "OrgDb")
orgdb <- query(orgs, "Homo sapiens")[[1]]
```

We can see the *OrgDb* bundles a lot of information from various
sources (Entrez, GO, KEGG, UCSC, Ensembl, Uniprot).

```{r}
orgdb
```

The `columns` function tells us what identifiers are contained in the
database, where the elements of the database are genes or associated proteins.

```{r}
columns(orgdb)
```

I typically like to
use [SFTPB](https://gtexportal.org/home/gene/SFTPB) as an example
gene. If you follow the link you can see it's expression profile in
humans across various tissues. It's a gene that is almost exclusively
expressed in lung, and it plays a role in producing *surfactant*,
which reduces the surface tension in the alveoli in the lung. The lung
is like a big balloon, but with many small sacks called alveoli, which
produce a massive amount of surface area in a small volume. Without
*surfactant*, we wouldn't have the muscle power to expand our lungs.

We can look up more information about SFTPB using `select`. We provide
a key, which is the gene symbol, and ask for the gene name, a column
of `orgdb`:

```{r}
select(orgdb, keys="SFTPB", columns="GENENAME", keytype="SYMBOL")
```

We can find out what roles a gene play, by looking up the *Gene
Ontology* terms, or GO terms:

```{r}
go <- select(orgdb, "SFTPB", "GO", "SYMBOL")
go
```

These are not informative alone, just, e.g. `GO:0001664`.
Let's restict to the terms that represent "biological processes":

```{r}
go <- go[go$ONTOLOGY == "BP",] # biological processes
go
```

Now we need to look up what the GO terms mean, using a separate
database, called `GO.db`:

```{r}
library(GO.db)
columns(GO.db)
go2 <- select(GO.db, go$GO, c("TERM","DEFINITION"), "GOID")
go2$TERM
```

As expected, one of the terms has to do with *respiratory gaseous
exchange*. Also, the surfactant is made up of many lipids, so the
first term is not surprising. Sometimes very generic categories show
up here, so it's relevant to know how many genes a GO term is
associated with. Here we will use the *Ensembl* IDs for genes.

```{r}
sapply(go$GO, function(term) nrow(select(orgdb, term, "ENSEMBL", "GO")))
```

Another gene I often use as an example is the *glucocorticoid
receptor*, which is the protein that responds to the hormone
*cortisol*, the "stress hormone". We can repeat these lines of code
with the gene symbol for glucocorticoid receptor (see the alias "GR"):

```{r}
select(orgdb, "NR3C1", "GENENAME", "SYMBOL")
select(orgdb, "NR3C1", "ALIAS", "SYMBOL")
go <- select(orgdb, "NR3C1", "GO", "SYMBOL")
go <- go[go$ONTOLOGY == "BP",] # biological processes
go2 <- select(GO.db, go$GO, c("TERM","DEFINITION"), "GOID")
go2$TERM
```
