---
title: "Genomic ranges: GRanges"
author: "[Michael Love](http://mikelove.github.io)"
output: html_document
editor_options: 
  markdown: 
    wrap: 80
---

Here we show off the way we specify genomic ranges in Bioconductor, using the
*GenomicRanges* package and the *GRanges* class.

A *GRanges* object specifies both the basepairs involved in the range (these are
integer ranges), the strand (+ or -), and the chromosome. Chromosomes in
Bioconductor are called "sequences" (because sometimes we may want to specify
ranges of non-chromosomes, for example viral DNA, un-incorporated contigs,
etc.).

# Interval ranges

Let's start with a set of interval ranges, an *IRanges* object.

```{r message=FALSE}
library(GenomicRanges)
ir <- IRanges(start=100 + 1:5, width=10)
ir
```

We could also specify the same ranges by setting the *end*:

```{r}
ir <- IRanges(start=100 + 1:5, end=100 + 10:14)
ir
```

We can make a plot of these ranges. It's a bit different than a typical plot,
because I want to point out that we have *discrete* locations in the genome (so
the position isn't defined over $\mathbb{R}$ but over $\mathbb{N}$). So I've
shifted the positions of the labels by 0.5 (see the `axis` command). Another
detail to notice is that the ranges are *inclusive*. So the first range, from
101 to 110 *includes* the 110 basepair. So I've added +1 when I draw the arrows
to show it includes the last basepair.

(Note that the convention used in Bioconductor to define interval ranges is not
universal. The BED convention is that the range does **not** include the start
position of the range. So 101-110 in BED format would not include the 101th
basepair.)

```{r message=FALSE}
library(rafalib)
nullplot(start(range(ir)), end(range(ir))+1, 0, 6, xaxt="n")
abline(v=100:115, col="grey")
axis(1, at=100:115 + .5, labels=100:115) 
arrows(start(ir),seq_along(ir),end(ir)+1,seq_along(ir),
       lwd=3,code=3,angle=90,length=.05)
```

# Genomic ranges

So far we've just shown interval ranges. To specify a genomic range, we also
will come up with the *strand* of the ranges (+ or -) and the chromosome.

```{r}
st <- rep(c("+","-"),c(3,2))
st
```

```{r}
gr <- GRanges(seqnames="chr1", ranges=ir, strand=st)
gr
```

Note that *GRanges* objects, when printed, also have a note about the genome.
Here it says that we have an unspecified genome, and there is no information
about the `seqlengths`, the lengths of the chromosomes. This information
(basically in what universe do the ranges live) is important to track along with
the ranges themselves, in case the user asks to, e.g. move all ranges 1 million
basepairs to the right. This is not necessarily possible if we have a range that
ends less than 1 million basepairs from the end of a chromosome. So the
*GRanges* object really needs to also contain details about the genome itself.

Here we show some basic manipulations of *GRanges*:

```{r}
gr + 10 # growing by 10 on either side
shift(gr, 10) # shifting by 10
resize(gr, width=1) # 1 bp from the start (+) or end (-)
flank(gr, width=2, both=TRUE) # +/- 2 bp from " or "
```

Note that the last two commands, `resize` and `flank`, gave back ranges relative
to the *start* of the ranges with positive strand, and to the *end* of the
ranges with negative strand. This is particular useful for working with genes,
in which we often care about the region surrounding the *transcription start
site* or TSS of a gene. The TSS will be the left-most basepair of a gene on the
positive strand, and the right-most basepair of a gene on the negative strand.

These functions would usually give us more information, but in this case the
information was not specified:

```{r}
seqnames(gr)
seqinfo(gr)
```

Instead of looking at a toy *GRanges*, let's download some actual genomic
ranges, the set of human genes, as annotated by the [Ensembl
project](http://www.ensembl.org/), version 86.

This data is easily accessible in Bioconductor using the *ensembldb* package,
and the specific Ensembl database package from Bioconductor.

This says there are about 60 thousand genes (according to Ensembl).

```{r message=FALSE}
library(ensembldb)
library(EnsDb.Hsapiens.v86) # this pkg is about 75 Mb
edb <- EnsDb.Hsapiens.v86
g <- genes(edb)
length(g)
```

If you try to print `g` it will just show you the beginning and end:

```{r}
g
```

You can take a look at a single gene:

```{r}
g[1]
```

To pull out the gene names you use a `$`. This would give us a 60 thousand
length character vector, so we just show the first 5.

```{r}
g$gene_name[1:5]
```

We can look at the distribution of genes by chromosome. The `seqnames` returns a
*factor-Rle* where *Rle* stands for *run length encoding*. This is a way to
compress a long vector which may contain repeated values in *runs*.

```{r}
seqnames(g)
```

The object `g` has information about genes on the standard chromosomes, as well
as genes or *haplotypes* of genes on *haplotype chromosomes*.

```{r}
head(seqlevels(g),100)
```

Let's subset to just the "standard" chromosomes:

```{r}
g <- keepStandardChromosomes(g, pruning.mode="coarse")
```

The following functions tell you the names and lengths of chromosomes, as well
as the name of the genome (repeated for each chromosome).

```{r}
seqlevels(g)
seqinfo(g)
genome(g)
```

We can get the number of genes on each chromosome:

```{r}
table(seqnames(g))
```

Note that MT (mitochondrial DNA) has many more genes than the other chromosomes,
per basepair:

```{r}
tab <- table(seqnames(g))
tab
lens <- seqlengths(seqinfo(g))
lens
all.equal(names(tab), names(lens))
barplot(tab/lens)
```

If we remove MT, we can see the range better:

```{r, fig.height=7}
barplot((tab/lens)[-which(names(tab) == "MT")], horiz=TRUE, las=1)
abline(v=(1:5)*1e-5, col=rgb(0,0,0,.5))
```

So roughly 1 gene per 50,000 basepairs, averaging per chromosome.

```{r}
mean((lens/tab)[-which(names(tab) == "MT")])
```

# Finding overlaps

Finally, I want to introduce some functionality we will use a couple times later
in the course: finding overlaps of sets of genomic ranges. One scenario is that
we have two sets and we want to know which ranges from the two sets overlap each
other. We might make a *query* set and ask which genes are overlapping the set
in our query.

```{r}
query <- GRanges(1, IRanges((101:110)*1e6 + 1,width=1e6))
query
```

We can get the numeric indices of the overlaps using `findOverlaps`. Note that
we can turn this resulting object into a data.frame, for further processing.

```{r}
fo <- findOverlaps(query, g)
head(fo)
tail(fo)
# to count the overlaps for each range in query:
table(factor(from(fo), seq_along(query)))
```

Equivalently, we can use the `countOverlaps` function for the last command
above:

```{r}
countOverlaps(query, g)
```

We can convert the Hits object to a data.frame:

```{r}
fo <- as.data.frame(fo)
```

We can also find out which genes overlap which other genes, by just providing a
single set of ranges to `findOverlaps`. There are *many* overlapping genes in
the human genome, though you should note that we obtain at least two rows in the
resulting *Hits* object for every overlap, and so we can filter this object down
by insisting that one index be strictly less than the other.

```{r}
fo <- findOverlaps(g)
head(fo)
fo <- as.data.frame(fo)
fo <- fo[fo[,1] < fo[,2],]
head(fo)
nrow(fo)
```

Earlier, when we defined *query*, we didn't consider strand, and so the strand
of the ranges was defined as `*`. This means that we get overlaps of any genes
whether + or -. But when we ask whether the genes overlap each other, by default
`findOverlaps` will not count an overlap if the ranges are on different strand.
To ask how many genes overlap each other ignoring strand, we need to specify
`ignore.strand=TRUE` (which produces more overlaps).

```{r}
fo <- findOverlaps(g, ignore.strand=TRUE)
fo <- as.data.frame(fo)
fo <- fo[fo[,1] < fo[,2],]
nrow(fo)
```

If we just wanted to subset to the genes which overlap a given range, we can use
`overlapsAny`:

```{r}
g[overlapsAny(g, query[1])]
```

This is equivalent to the following:

```{r}
g[g %over% query[1]]
```

# plyranges

As with *SummarizedExperiment*, there is a "tidy" way to work with *GRanges*,
called *plyranges*. For more examples of this type of analysis, take a look at
some tutorial content I've been working on here:

<https://nullranges.github.io/tidy-ranges-tutorial>

```{r message=FALSE}
library(plyranges)
g %>% 
  group_by(gene_biotype) %>%
  summarize(width = mean(width))
```

Count overlaps of genes in the query ranges:

```{r}
query %>%
  mutate(num_genes = count_overlaps(., g))
```

We can also do the same types of operations as before, with "base *GRanges*":

```{r}
g %>%
  slice(1:5) %>%
  select(symbol) %>%
  anchor_5p() %>%
  mutate(width=1) # same as resize(..., width=1)
```

Plot genes per chromosome:

```{r}
library(tibble)
library(ggplot2)
g %>% 
  group_by(seqnames) %>%
  summarize(num_genes = n()) %>%
  as_tibble() %>%
  ggplot(aes(seqnames, num_genes)) +
  geom_col()
```

```{r}
sessionInfo()
```
