---
title: "Hierarchical modeling of variance"
author: "[Michael Love](http://mikelove.github.io)"
output: html_document
---

This week we will focus on hierarchical models as they are used in
genomics, and particularly hierarchical models for estimates of
variance. A good review of this topic is:

[Analyzing ’omics data using hierarchical models](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2904972/) 
by Hongkai Ji and X Shirley Liu

Hierarchical models have proven very useful in genomics, when we often
have precious few biological replicates to assess within-group
variance. Often experiments themselves, as well as the technical costs
of sequencing DNA, mean that only 3-5 samples might be generated for
each biological condition, although there may be many (e.g. dozens) of
conditions. Ideally, more samples would be generated
for definitive estimation of differences relative to biological
variability in each condition. This all depends greatly on what the
treatments are doing, and the relationship of the replicates to each
other -- are they mice in a controlled setting with the same genetic
background, or human donors in a clinic, etc.?

The point of the hierarchical model is summarized nicely in 
[Figure 1](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2904972/figure/F1/)
from the paper linked above. The key idea is that, we only observe a
few samples and therefore, we can potentially have very bad estimates
of the variance for some genes, either because the samples were
observe happened to be close to each other (under-estimation) or too
spread apart (over-estimation), and do not represent the biological
variability we would see if we had generated more
replicates. Hierarchical models, and an empirical Bayes procedure to
adjust the smallest and largest estimates of variance towards the
*middle* of the distribution of sample variances, helps to avoid
large mistakes in inference when the estimated variances are used to
generate test statistics.

One of the most popular methods which made use of the
hierarchical model for variance estimation is 
[limma](http://bioinf.wehi.edu.au/limma/), which was originally
designed for microarray (continuous valued), but has been extended
recently to work with sequencing data (counts). 
In this document, I will show the practical effect of running the
`eBayes` (empirical Bayes) function in limma, and how this modifies
the variance estimates.

```{r echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r message=FALSE}
library(curatedBladderData)
library(affy)
data(GSE13507_eset)
e <- GSE13507_eset
```

```{r}
e <- e[,which(e$summarystage == "invasive")]
dim(e)
boxplot(exprs(e)[,1:10], range=0, main="samples")
boxplot(t(exprs(e)[1:10,]), range=0, main="genes")
```

```{r message=FALSE}
library(matrixStats)
rm <- rowMeans(exprs(e))
rv <- rowVars(exprs(e))
plot(rm, sqrt(rv), cex=.5, col=rgb(0,0,0,.4))
```

```{r}
e <- e[rm > 8,]
rv <- rowVars(exprs(e))
hist(sqrt(rv),breaks=50,col="grey")
```

```{r}
hc <- hclust(dist(t(exprs(e)[order(rv,decreasing=TRUE)[1:1000],]))) 
plot(hc)
```

```{r fig.width=6, fig.height=6}
m <- 5
sample.idx <- sample(ncol(e), m)
e.sub <- e[,1:m]
rv.sub <- rowVars(exprs(e.sub))
plot(sqrt(rv), sqrt(rv.sub), cex=.5, col=rgb(0,0,0,.5),
     xlab="sample SD full data", ylab="sample SD subset")
abline(0,1)
```

```{r message=FALSE}
library(limma)
design <- model.matrix(~1, data.frame(row.names=1:m))
fit <- lmFit(exprs(e.sub), design)
fit <- eBayes(fit)
```

```{r, fig.width=6, fig.height=6}
plot(rv.sub, fit$s2.post, xlim=c(0,1), ylim=c(0,1),
     xlab="sample var", ylab="eBayes var")
abline(0,1)
abline(v=fit$s2.prior, h=fit$s2.prior)
```

```{r message=FALSE, fig.width=5, fig.height=5}
library(rafalib)
nullplot(0,1,0,2,xaxt="n")
axis(1, c(0,1), c("before","after"))
n <- 100
idx <- sample(nrow(e),n)
segments(rep(0,n), rv.sub[idx], rep(1,n), fit$s2.post[idx], col=rgb(0,0,0,.4))
abline(h=fit$s2.prior, col="dodgerblue", lwd=5)
```

```{r fig.width=5, fig.height=5}
nullplot(0,1,0,12,xaxt="n")
axis(1, c(0,1), c("before","after"))
n <- 100
idx <- order(rv.sub, decreasing=TRUE)[1:n]
segments(rep(0,n), rv.sub[idx], rep(1,n), fit$s2.post[idx], col=rgb(0,0,0,.4))
abline(h=fit$s2.prior, col="dodgerblue", lwd=5)
```

```{r}
true <- sqrt(rv)
samp.sd <- sqrt(rv.sub)
ebayes.sd <- sqrt(fit$s2.post)
sqrt(mean((samp.sd - true)^2))
sqrt(mean((ebayes.sd - true)^2))
median(abs(samp.sd - true))
median(abs(ebayes.sd - true))
```

```{r, fig.width=8, fig.height=5}
par(mfrow=c(1,2),mar=c(4,2,2,1))
plot(samp.sd, samp.sd - true, col=rgb(0,0,0,.4), xlim=c(0,4), ylim=c(-1.5,1.5),
     main="residual over estimate")
abline(h=-1:1,v=.5,lty=2,col=rgb(1,0,0,.5))
plot(ebayes.sd, ebayes.sd - true, col=rgb(0,0,0,.4), xlim=c(0,4), ylim=c(-1.5,1.5),
     main="residual over estimate")
abline(h=-1:1,v=.5,lty=2,col=rgb(1,0,0,.5))
```

```{r}
plot(true, samp.sd, type="n")
alpha <- ifelse(true > 1.5, .5, ifelse(true > 1, .2, .1))
arrows(true, samp.sd, true, ebayes.sd, col=rgb(0,0,0,alpha), length=.1)
abline(0,1)
```
