---
title: "Hierarchical modeling of variance"
author: "[Michael Love](http://mikelove.github.io)"
output: html_document
---

This week we will focus on hierarchical models as they are used in
genomics, and particularly hierarchical models for estimates of
variance. A good review of this topic is:

[Analyzing ’omics data using hierarchical models](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2904972/) 
by Hongkai Ji and X Shirley Liu

Hierarchical models have proven very useful in genomics, when we often
have precious few biological replicates to assess within-group
variance. Often experiments themselves, as well as the technical costs
of sequencing DNA, mean that only 3-5 samples might be generated for
each biological condition, although there may be many (e.g. dozens) of
conditions. Ideally, more samples would be generated
for definitive estimation of differences relative to biological
variability in each condition. This all depends greatly on what the
treatments are doing, and the relationship of the replicates to each
other -- are they mice in a controlled setting with the same genetic
background, or human donors in a clinic, etc.?

The point of the hierarchical model is summarized nicely in 
[Figure 1](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2904972/figure/F1/)
from the paper linked above. The key idea is that, we only observe a
few samples and therefore, we can potentially have very bad estimates
of the variance for some genes, either because the samples were
observe happened to be close to each other (under-estimation) or too
spread apart (over-estimation), and do not represent the biological
variability we would see if we had generated more
replicates. Hierarchical models, and an empirical Bayes procedure to
adjust the smallest and largest estimates of variance towards the
*middle* of the distribution of sample variances, helps to avoid
large mistakes in inference when the estimated variances are used to
generate test statistics. This is sometimes referred to as *moderation*
of estimates, and the estimates themselves are sometimes referred to
as *shrinkage estimators*. 

One of the most popular methods which made use of the
hierarchical model for variance estimation is 
[limma](http://bioinf.wehi.edu.au/limma/), which was originally
designed for microarray (continuous valued), but has been extended
recently to work with sequencing data (counts). 
Other methods for count data, such as DESeq2 and edgeR, likewise adopt
a shrinkage based method similar to the one shown here, but for the
dispersion parameter of a count distribution.
In this document, I will show the practical effect of running the
`eBayes` (empirical Bayes) function in limma, and how this modifies
the variance estimates.

I start by loading the curatedBladderData package, which has a number
of *ExpressionSet* objects with gene expression from patients with
bladder cancer. We load one of the dataset, and examine the experiment
data.

```{r echo=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

```{r message=FALSE}
library(curatedBladderData)
library(affy)
data(GSE13507_eset)
e <- GSE13507_eset
str(experimentData(e))
```
We will subset to a set of samples with the same staging. Our goal
will be to estimate the variance of each gene using a subset of
samples that is typical of a small experiment (n=5 vs 5), and compare
to the variance estimate we get using all of the samples.

```{r}
table(e$summarystage)
e <- e[,which(e$summarystage == "superficial")]
dim(e)
boxplot(exprs(e), range=0, main="samples")
boxplot(t(exprs(e)[1:10,]), range=0, main="genes")
```

A plot of the variance over the mean for each gene. For this
demonstration, we will remove the genes with low mean, where the
variances approach 0. 

```{r message=FALSE}
library(matrixStats)
rm <- rowMeans(exprs(e))
rv <- rowVars(exprs(e))
plot(rm, sqrt(rv), cex=.5, col=rgb(0,0,0,.4))
```

A histogram of the variances for each gene:

```{r}
e <- e[rm > 8,]
rv <- rowVars(exprs(e))
hist(sqrt(rv),breaks=50,col="grey")
```

Finally, we perform a quick check to see that there are not large
clusters in the data. 

```{r}
hc <- hclust(dist(t(exprs(e)[order(rv,decreasing=TRUE)[1:1000],]))) 
plot(hc)
```

There seems to be a single outlier, which we remove.

```{r}
e <- e[,-which(colnames(e)=="GSM340606")]
```

Let's recalculate the sample variance of each row, for use later:

```{r}
rv <- rowVars(exprs(e))
```

We take a random sub-sample of just 10 to compare with the full
sample set. Plotting the sample variance of the subset against the
full sample set shows some correspondence, but many above and below
the line.

```{r fig.width=6, fig.height=6}
n <- 10
set.seed(1)
sample.idx <- sample(ncol(e), n)
e.sub <- e[,sample.idx]
rv.sub <- rowVars(exprs(e.sub))
plot(sqrt(rv), sqrt(rv.sub), cex=.5, col=rgb(0,0,0,.5),
     xlab="sample SD full data", ylab="sample SD subset")
abline(0,1)
```

# Empirical Bayes shrinkage estimator

The limma package has a function `eBayes` which takes in sample
variance estimates, calculates the parameters of a prior distribution
by fitting it to the observed data, and then produces posterior
estimates for the sample variance. Plotting the posterior estimates
against the standard sample variances, you can see they have shifted
toward a central value, which is `s2.prior`.

```{r message=FALSE}
library(limma)
design <- model.matrix(~1, data.frame(row.names=1:n))
fit <- lmFit(exprs(e.sub), design)
fit <- eBayes(fit)
```

```{r, fig.width=5, fig.height=5}
plot(rv.sub, fit$s2.post, xlim=c(0,1), ylim=c(0,1),
     xlab="sample var", ylab="eBayes var")
abline(0,1)
abline(v=fit$s2.prior, h=fit$s2.prior)
```

Another way to visualize the posterior estimates is to plot the
original estimates on the left side, and the new estimates on the
right side, with segments connecting them. The blue line indicates the
middle of the prior distribution. Compare the following plot with the
second figure in 
[this paper](http://statweb.stanford.edu/~ckirby/brad/other/Article1977.pdf)
by Bradley Efron and Carl Morris on shrinkage estimators.

```{r message=FALSE, fig.width=5, fig.height=5}
library(rafalib)
nullplot(0,1,0,2,xaxt="n")
axis(1, c(0,1), c("before","after"))
n <- 100
idx <- sample(nrow(e),n)
segments(rep(0,n), rv.sub[idx], rep(1,n), fit$s2.post[idx], col=rgb(0,0,0,.4))
abline(h=fit$s2.prior, col="dodgerblue", lwd=5)
```

Repeating the above plot by seeing that the highest estimates of
sample variance are "shrunk" significantly toward the middle.

```{r fig.width=5, fig.height=5}
nullplot(0,1,0,12,xaxt="n")
axis(1, c(0,1), c("before","after"))
n <- 100
idx <- order(rv.sub, decreasing=TRUE)[1:n]
segments(rep(0,n), rv.sub[idx], rep(1,n), fit$s2.post[idx], col=rgb(0,0,0,.4))
abline(h=fit$s2.prior, col="dodgerblue", lwd=5)
```

We can show that we have reduced both the root mean squared error, as
well as the median absolute error with our new estimators, compared to
"true": the sample variance using all the samples.

```{r}
true <- sqrt(rv)
samp.sd <- sqrt(rv.sub)
ebayes.sd <- sqrt(fit$s2.post)
sqrt(mean((samp.sd - true)^2))
sqrt(mean((ebayes.sd - true)^2))
median(abs(samp.sd - true))
median(abs(ebayes.sd - true))
```

It helps to zoom into individual genes, to see how we have reduced the
large errors comparing our estimate on a small
sub-sample to the full dataset. Two things to note: 
(i) when the estimates are low, they tend to be too low, and when the
estimates are high, they tend to be too high. This phenomenon is
sometimes referred to as "winner's curse".
(ii) most of the posterior variance estimates are close to 0.5, so we
have reduced the number of genes on the left side, where the variance
estimate is very low, and underestimated. These genes would lead to
false positives, by inflating the *t* statistic. Similarly, using the
posterior, there are fewer genes with large overestimates of the
variance.

```{r, fig.width=7, fig.height=4}
par(mfrow=c(1,2),mar=c(4,2,2,1))
plot(samp.sd, samp.sd - true, col=rgb(0,0,0,.4), xlim=c(0,4), ylim=c(-1.5,1.5),
     main="'error' over estimate")
abline(h=-1:1,v=.5,lty=2,col=rgb(1,0,0,.5))
plot(ebayes.sd, ebayes.sd - true, col=rgb(0,0,0,.4), xlim=c(0,4), ylim=c(-1.5,1.5),
     main="'error' over estimate")
abline(h=-1:1,v=.5,lty=2,col=rgb(1,0,0,.5))
```

One final plot to show the change is to draw arrows when the two
estimates disagree, over the "true" sample variance from the full
dataset. Note that, the posterior estimates are not perfect, and in
some cases we have moved the estimates in the wrong
direction. However, overall, we have improved more estimates than we
have made worse.


```{r, fig.width=6, fig.height=6}
plot(true, samp.sd, type="n", ylab="sample SD --> eBayes SD")
idx <- abs(samp.sd - ebayes.sd) > .1
arrows(true[idx], samp.sd[idx],
       true[idx], ebayes.sd[idx],
       col=rgb(0,0,0,0.3), length=.1)
abline(0,1)
```
