---
title: "Homework 4 - Data modeling: EM and HMM"
author: "your name here"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
---

# Question 1 - EM for cell types (12 pts)

Suppose we measure 10 proteins on the surface of many cells (the data
$X$). The cells are of a fixed (6) number of underlying cell types
(the latent variable $Z$). Each cell type has a pattern of surface
protein abundance (the model parameters $\theta$). We will use a
simplified model and EM to estimate $\theta$ and a posterior
distribution for $Z$.

```{r}
nproteins <- 10
ncelltypes <- 6
set.seed(123)
# this will be the true model ('theta')
cell_types <- matrix(rnorm(nproteins * ncelltypes), ncol=ncelltypes)
```

```{r}
palette("Okabe-Ito")
# plot of the protein abundance (y axis) for the 10 proteins (x axis)
matplot(cell_types, type="b", xlab="protein", ylab="abundance",
        main="protein abundance per cell type")
abline(h=0, col=rgb(0,0,0,.25))
```

The protein measurements $X$ for 1,000 cells are not identical to the
6 cell types, but have some variance/error:

```{r}
ncells <- 1000
# true 'z' used for simulation, but consider unobserved
z <- sample(ncelltypes, ncells, TRUE)
table(z)
x <- cell_types[,z] + matrix(rnorm(nproteins * ncells), 
                             ncol=ncells)
```

```{r}
matplot(x[,1:50], type="l", col=z[1:50], lty=1,
        xlab="protein", ylab="abundance",
        main="protein abundance per cell")
legend("topright", legend=1:6, col=1:6, lty=1, inset=.05, lwd=2)
```

Simplifying assumptions: here we will estimate model parameters
$\theta = \{\vec{\mu}_1, ..., \vec{\mu}_6,\}$, the mean vectors for
the 6 cell types, where each is of length 10. You can assume the
variance of the data $X$ is given: $\sigma^2 = 1$, and does not need
to be estimated.

Part a: Write out in code the conditional distribution (or posterior)
of $Z$ given $X$ and $\theta^t$. You can write it out by computing
this for a randomly initialized $\theta^0$ (e.g. draw its values
$N(0,1)$), and/or by defining a function that takes $X$ and $\theta^t$
and returns the conditional distribution.

Part b: Perform EM for $\theta$ by maximizing the expectation of the
log likelihood with respect to this conditional distribution (or
posterior) from part a. Again, you can assume that the variance on
measurements for each protein is equal to 1. For maximization, you
don't need to analytically derive this, if you just want to refer to
the derivations seen in EM lecture for the mean of a mixture.

Part c: Examine $\hat{\theta}^{20}$, the estimates after the 20th
iteration. Does it correspond to the true $\theta$ ? Is each cell type
represented? Are they one-to-one? Also examine the posterior of $Z$
given $X$ and $\hat{\theta}^{20}$: was each cell appropriately
assigned to one of your estimated cell types? Were they one-to-one?
You should also try a couple of random iterations to see if your
conclusions generally hold (you don't need to show the repeated random
initiations in the code however).

# Question 2 - HMM for time series (12 pts)

Suppose we measure gene expression (here assume some normalized,
variance stabilized measurements) for 20 genes across a time course
with $T = 12$ time points. We are interested in finding a shift in the
expression during the time course. We will use an HMM, similar to the
one seen in class, to model change point(s) in the data, by seeing
when we observe a transition from one state with one mean value to
another state with another mean value. The key modification is that,
instead of observing one process as in the lecture note, here we
observe 20 of them at once.

Simplifying assumptions: we will assume that the 20 genes have been
standardized such that they all share the same mean when they are in
the same state. While the HMM could be made more complex to handle
differing mean value per gene per state, for this homework assume a
common mean (and variance) for all the genes, when they are in the
same state. Another simplifying assumption: instead of modeling the
variance, we will just assume $\sigma^2 = 1$ as in the first question,
and we will not include the variance as part of the estimation task.

Note: for this problem, we use some informative initialization, and
you are not asked to attempt multiple random initializations.

```{r}
# make some data with multiple genes, some kind of transition
set.seed(7)
ngenes <- 20
T <- 12
y_matrix <- matrix(rnorm(ngenes * T), ncol=T)
# the true transition: last four time points shift up by 0.7
y_matrix[,9:12] <- y_matrix[,9:12] + .7 
n <- 2 # just two states
```

```{r}
# need to reset plot par from before
par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
matplot(t(y_matrix[1:9,]), type="b")
```

```{r}
# initialize:
pi <- c(.95, .05) # equal starting probabilities
A <- matrix(1/T, nrow=n, ncol=n) # the transition matrix
diag(A) <- 11/T # high density on diagonal usually stay in same state
rowSums(A)
mu <- c(-.1, .1)
sigma <- rep(1, n) # oracle variance
# save initial parameters
pi0 <- pi
A0 <- A
mu0 <- mu
```

Part a: Write a function which performs one step of the EM for the
start probabilities `pi`, the transition matrix `A`, and the means
`mu`. To do so, you will need to modify the following from the
lecture: the function `B` to take a column vector, and the computation
for `alpha`, `beta`, and `xi` to deal with a matrix of observed data
instead of a single vector of data over time. You will also need to
change the update rule for `mu` so that it accounts for the sum over
genes (i.e. normalize the numerator by the number of genes).

Part b: Run the EM for 10 iterations and print the mean values at each
step. Print the final transition matrix. Do these values for the mean
and the transition matrix seem plausible as a good fit to the data?
How do you interpret the final values for $A_{2,1}$ and $A_{1,2}$?

Part c: Run the Viterbi algorithm and plot the matrix `V` over time
(e.g. you can use `matplot` to show the two states and their values
over time). Recall that $V_{i,t}$ gives the probability of the most
probable sequence up to *t* having *i* as its final state. What do you
observe? Did the HMM correctly identify the shift in the data across
all the genes?
