---
title: "Homework 6 - Hierarchical models / HMM"
author: "your name here"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
---

# Question 1 - compare limma's posterior variances to sample variances

We saw in `hierarchical.Rmd` how limma moderates its variance
estimates for each gene toward a middle value that is chosen by
looking at the sample variance estimates from all genes. In this
question we will explore how, on random subsets of the samples of
various size, the limma moderated variance estimate compares to the
standard sample variance estimate.

We will make the comparison on the scale of
**standard deviation**, and we will consider the sample standard
deviation over the full dataset as the *true* standard deviation of
the population. 

You should then calculate the root mean squared error (RMSE) of the
square root of the limma moderated variance on the subset to the
*true* SD and of the sample standard deviation on the subset to the
*true* SD.

The following code, from `hierarchical.Rmd` loads the expression data:

```{r message=FALSE}
library(curatedBladderData)
library(affy)
data(GSE13507_eset)
e <- GSE13507_eset # the microarray dataset
e <- e[,which(e$summarystage == "superficial")]
```

Calculating the row means and imposing a minimal expression filter:

```{r}
library(matrixStats)
rm <- rowMeans(exprs(e))
e <- e[rm > 8,]
# outlier:
e <- e[,-which(colnames(e)=="GSM340606")]
```

This is the per-gene sample standard deviation (over 102 samples) that
we will consider as the *true* standard deviation:

```{r}
gene_sd <- rowSds(exprs(e))
hist(gene_sd, breaks=40)
```

You should use the following vector of sample sizes, *n*, implying
that for each sample size from 3 to 20, you should do 5 random subsets
of the data. For each iteration, you should run limma (with design
`~1`)) and obtain the posterior variance estimate `s2.post`, which you
will then take the square root of, to obtain the posterior standard
deviation estimate.

You should also calculate the sample standard deviation for the random
subsets.

```{r}
ns <- rep(c(seq(3,20)),each=5)
table(ns)
```

You can plot the RMSE (comparing SD estimates) for both methods across
the iterations however you like, but ideally you would show the
individual RMSE’s for each iteration (so five points per sample size),
as well as the average or a smoothed estimate of the RMSE trend for
each estimator.

How would you describe the differences you see in the method
performance as *n* goes from 3 to 20?

# Question 2 - Use HMM to segment chromosome into isochores

In this last question, we will segment regions of a human chromosome
according to average GC content in windows of length
100kb. *Isochores* are defined as large regions of DNA that have
similar GC content, although GC content varies much more as we look
across a chromosome. Basically, we can see wavy patterns of average GC
content in the 100kb range. The isochores for human chromosomes are
described in this paper:

<http://genome.cshlp.org/content/16/4/536.full>

In particular see Figure 1:

<http://genome.cshlp.org/content/16/4/536/F1.expansion.html>

We will use a three state HMM to segment human chromosome 3 by its GC
content.

We start by building ranges of width 100kb that span from 1 to
198,200,000 on chromosome 3 (you should use
*BSgenome.Hsapiens.UCSC.hg38* that we used in the `bioc/strings.Rmd`
lecture note). I demonstrate below how to do this using `tileGenome`
from the *GenomicRanges* package.

You should end up with 1982 non-overlapping ranges. 

```{r}
library(GenomicRanges)
library(BSgenome.Hsapiens.UCSC.hg38)
head(seqlengths(Hsapiens))
tiles <- tileGenome(seqlengths(Hsapiens)[3],
                    tilewidth=100e3,
                    cut.last.tile.in.chrom=TRUE)
tiles <- tiles[width(tiles) == 100e3] # remove last tile
tiles
```

Get the DNA sequence of these ranges, and calculate the average GC
content (reminder from that lecture note: `letterFrequency(dna, "CG",
as.prob=TRUE)`).

For performing the segmentation, it's best to remove a few stray
ranges that have very low GC content, so use a filter of:

```{r eval=FALSE}
gc <- gc[gc > .3]
```

First, plot the GC content of the 1979 remaining ranges. You should
see a wavy pattern, reflecting the following quote from the paper
referenced above:

> genomes of warm-blooded vertebrates ... are mosaics of isochores,
> long (>>300 kb), compositionally fairly homogeneous regions that
> belong to a small number of families characterized by different
> average GC levels

Now, construct an HMM to segment the sequence. First initialize the
parameter values for $\theta = (\pi, A, B)$

For the initial transition matrix A, use a probability of .98 to stay
in the same state, and a probability of .01 to transition to either of
the other states. Use initial mu values of [.3, .4, .5] and sigma of
.05. 

**Note:** we will not use Baum-Welch to update `sigma` in this
case, as it leads to non-identifiability of solutions. We will only
update `A`, `pi`, and `mu`.

Update the parameters with **only one iteration of Baum-Welch / EM**.

Now with updated parameters, find the Viterbi path. Plot the Viterbi
sequence below a plot of the data. How many segments ("runs" of the
same state) do you obtain? (You can use `rle` to calculate how many
runs of the same state you obtain.)

Finally, plot the data (the $y_t$), and color the points according to
the Viterbi path. On top of the colored points draw the fitted mean
values for the three states (e.g. as in lecture note). Qualitatively
compare the fitted mean values you find with the modes of the
histogram from this figure in the paper:

<http://genome.cshlp.org/content/16/4/536/F6.expansion.html>

