---
title: "Homework 8 - Genomic signal processing with HMM"
author: "your name here"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
---

# Question 1 - Add a fourth state to the HMM

In `hmm.Rmd` we segmented arrayCGH data for a single sample using 3
hidden states of an HMM. Here we will add a fourth state, to deal with
the different apparent mean value of the two duplications.
Start by loading the arrayCGH data as we did in `hmm.Rmd`. Modify the
HMM that we had in class so it has a fourth state to accommodate the
two levels of the duplications. So the four states will be [deleted,
normal, duplicated type 1, duplicated type 2]. 

You should change the initial transition matrix `A`, so that the
non-normal copy states have probability of .99 to stay in the same
state, and .01 to go back to normal state (state 2). The normal state
(state 2) will have probability .97 to stay normal, and probability of
.01 to go to each of the non-normal states.

Use the following as the initial vector for `mu`:

```{r}
mu <- c(-.5,0,.3,.8) 
```

The one major change from the code in `hmm.Rmd` is that we will **first**
update the parameters using one iteration of the Baum-Welch algorithm
(EM algorithm for HMM), and **then** calculate the Viterbi path using
the updated parameters. This is a more realistic series of
operations. 

Finally, draw the Viterbi path using colored points (as in
`hmm.Rmd`). Draw the final Viterbi path once showing with colored
horizontal lines the initial `mu` values, and the final Viterbi path
again with lines showing the final `mu` values. (You don't have to
calculate the Viterbi path twice, in the plot we just change the
location of the horizontal lines.)

Does the fourth state help in identifying the two duplications as
being in different state?

# Bonus: run DNAcopy's segmentation algorithm

Run DNAcopy on `smo` and qualitatively compare with the Viterbi path
we found above. You can find code for running the segmentation
algorithm implemented in *DNAcopy* by typing
`vignette("DNAcopy")`. Does DNAcopy find more or less "non-normal"
segments than our HMM's Viterbi path? How do the fitted means compare
qualitatively?

# Question 2

In this last question, we will segment regions of a human chromosome
according to average GC content in windows of length
100kb. *Isochores* are defined as large regions of DNA that have
similar GC content, although GC content varies much more as we look
across a chromosome. Basically, we can see wavy patterns of average GC
content in the 100kb range. The isochores for human chromosomes are
described in this paper:

<http://genome.cshlp.org/content/16/4/536.full>

In particular see Figure 1:

<http://genome.cshlp.org/content/16/4/536/F1.expansion.html>

We will use a three state HMM to segment human chromosome 3.

Build ranges of width 100kb that span from 1 to 198,000,000 on
chromosome 3 (you should use *BSgenome.Hsapiens.UCSC.hg38* that we
used in the `bioc/strings.Rmd` notes). So you will end up with 1980
non-overlapping ranges. Get the DNA sequence of these ranges, and
calculate the average GC content. For performing the segmentation,
it's best to remove a few stray ranges that have very low GC content, so
use a filter of:

```{r eval=FALSE}
gc <- gc[gc > .3]
```

First, plot the GC content of the 1977 remaining ranges.

For the initial transition matrix A, use a probability of .98 to stay
in the same state, and a probability of .01 to transition to either of
the other states. Use initial mu values of [.3, .4, .5] and sigma of
.05. **Note:** we will not use Baum-Welch to update sigma in this
case, as it leads to non-identifiability of solutions. We will only
update `A`, `pi`, and `mu`. 

After updating the parameters with one iteration of Baum-Welch, find
the Viterbi path. Plot the Viterbi sequence below a plot of the
data. How many segments ("runs" of the same state) do you obtain?

Finally, plot the data and color the points according to the Viterbi
path. On top of the colored points draw the fitted mean values for the
three states. Qualitatively compare the fitted mean values you find
with the modes of the histogram from this figure in the paper:

<http://genome.cshlp.org/content/16/4/536/F6.expansion.html>
