---
title: "Classification across three brain regions"
author: "[Michael Love](http://mikelove.github.io)"
format:
  html:
    embed-resources: true
---

```{r message=FALSE}
library(here)
library(SummarizedExperiment)
library(magrittr)
library(matrixStats)
library(ggplot2)
library(GGally)
source(here("bioc","my_scale_counts.R"))
```

Load and scale data.

```{r}
load(here("eda","brain.rda"))
se <- my_scale_counts(rse_gene)
cts <- assay(se)
condition <- se$smtsd
condition %<>% (function(x) sub("Brain - (.*)", "\\1", x))
condition %<>% (function(x) sub("Spinal cord \\(.*\\)", "Spinal cord", x))
condition %<>% factor
```

Filter genes with low total counts.

```{r}
cts <- cts[rowSums(cts) > 100,]
dim(cts)
```

Identify Spinal cord, Hypothalamus, and Hippocampus samples.

```{r}
sc_idx    <- grep("Spinal cord",  as.character(condition))
hypo_idx  <- grep("Hypothalamus", as.character(condition))
hippo_idx <- grep("Hippocampus",  as.character(condition))
cat("Spinal cord:", length(sc_idx),
    "  Hypothalamus:", length(hypo_idx),
    "  Hippocampus:", length(hippo_idx), "\n")
```

Subset to the three regions, log-transform, and select the top 500 most variable genes.

```{r}
idx     <- c(sc_idx, hypo_idx, hippo_idx)
logcts  <- log10(cts[, idx] + 1)
y       <- droplevels(condition[idx])

rv <- rowVars(logcts)
o  <- order(rv, decreasing = TRUE)[1:500]
x  <- logcts[o, ]
dim(x)
```

Run PCA on the 500-gene log-count matrix (samples as rows).

```{r}
pc    <- prcomp(t(x))
x_pca <- pc$x
```

Plot all pairs of PC1 through PC4, colored by brain region.
Hippocampus is shown in goldenrod, Hypothalamus in steelblue, and Spinal cord in tomato.

```{r fig.width=8, fig.height=7, message=FALSE}
region_colors <- c("Hippocampus" = "goldenrod",
                   "Hypothalamus" = "steelblue",
                   "Spinal cord"  = "tomato")
dat <- data.frame(x_pca[, 1:4], condition = y)
ggpairs(dat,
        columns = 1:4,
        aes(color = condition, alpha = 0.4),
        upper = list(continuous = "points"),
        lower = list(continuous = "points"),
        diag  = list(continuous = "densityDiag")) +
  scale_color_manual(values = region_colors) +
  scale_fill_manual(values = region_colors) +
  theme_bw()
```

Some more information about the genes:

```{r}
rngs <- rowRanges(se)
gene_dat <- data.frame(
  row.names = names(rngs),
  chrom = seqnames(rngs),
  strand = strand(rngs),
  width = width(rngs),
  symbol = sapply(rngs$symbol, `[`, 1)
)
head(gene_dat)
if (!identical(rownames(gene_dat), rownames(x)))
  gene_dat <- gene_dat[rownames(x),,drop=FALSE]
```

```{r}
sessionInfo()
```
