---
title: "Interpolation between brain regions"
author: "[Michael Love](http://mikelove.github.io)"
format:
  html:
    embed-resources: true
---

```{r message=FALSE}
library(here)
library(SummarizedExperiment)
library(magrittr)
library(matrixStats)
library(ggplot2)
source(here("bioc","my_scale_counts.R"))
```

Load and scale data.

```{r}
load(here("eda","brain.rda"))
se <- my_scale_counts(rse_gene)
cts <- assay(se)
condition <- se$smtsd
condition %<>% (function(x) sub("Brain - (.*)", "\\1", x))
condition %<>% factor
```

Filter genes with low total counts.

```{r}
cts <- cts[rowSums(cts) > 100,]
dim(cts)
```

Identify Spinal cord and Hypothalamus samples.

```{r}
sc_idx <- grep("Spinal cord", as.character(condition))
hypo_idx <- grep("Hypothalamus", as.character(condition))
cat("Spinal cord:", length(sc_idx), "  Hypothalamus:", length(hypo_idx), "\n")
```

Pick one sample from each region and create 10 pseudosamples by linear interpolation on the log count scale. A weight of 0 gives the Hypothalamus sample, weight 1 gives the Spinal cord sample.

```{r}
sc_log   <- log10(cts[, "SRR1402882"] + 1)
hypo_log <- log10(cts[, "SRR1413454"] + 1)

weights <- seq(0, 1, length.out = 10)
pseudo  <- sapply(weights, function(w) w * sc_log + (1 - w) * hypo_log)
colnames(pseudo) <- paste0("pseudo_", seq_len(10))
```

Combine with all Spinal cord and Hypothalamus samples, log-transform the real samples, and run PCA on the top 500 most variable genes.

```{r}
logcts <- cbind(log10(cts[, sc_idx] + 1), log10(cts[, hypo_idx] + 1), pseudo)
group <- c(
  rep("Spinal cord", length(sc_idx)),
  rep("Hypothalamus", length(hypo_idx)),
  rep("Pseudosample", 10)
)

rv <- rowVars(logcts)
o  <- order(rv, decreasing = TRUE)[1:500]
pc <- prcomp(t(logcts[o,]))
```

Pseudosamples are colored by interpolation weight (0 = Hypothalamus, 1 = Spinal cord). Real samples are shown as filled circles.

```{r fig.width=7, fig.height=5}
dat      <- data.frame(pc$x[, 1:2], group = group)
dat_real <- dat[dat$group != "Pseudosample",]
dat_pseudo <- dat[dat$group == "Pseudosample",]
dat_pseudo$weight <- weights

ggplot() +
  geom_point(data = dat_pseudo,
             aes(PC1, PC2, color = weight), shape = 18, size = 4, alpha = 0.9) +
  scale_color_gradient(low = "tomato", high = "steelblue",
                       name = "interp.\nweight") +
  geom_point(data = dat_real,
             aes(PC1, PC2, fill = group), shape = 21, size = 2.5, stroke = 0.5) +
  scale_fill_manual(values = c("Spinal cord" = "steelblue",
                               "Hypothalamus" = "tomato")) +
  theme_bw()
```

```{r}
sessionInfo()
```
