---
title: "Brain RNA EDA Exercises"
author: ""
format:
  html:
    embed-resources: true
execute:
  eval: false
---

## Setup

Run these chunks first to build the objects used throughout the exercises.
This mirrors the setup in `brain_RNA.qmd`.

```{r}
library(here)
library(SummarizedExperiment)
library(magrittr)
library(matrixStats)
library(ggplot2)
```

```{r}
url <- "http://duffel.rail.bio/recount/SRP012682/rse_gene_brain.Rdata"
file <- here("eda", "brain.rda")
if (!file.exists(file)) download.file(url, file)
load(file)
```

```{r}
source(here("bioc", "my_scale_counts.R"))
se <- my_scale_counts(rse_gene)
```

```{r}
cts <- assay(se)
condition <- se$smtsd
condition %<>% (function(x) sub("Brain - (.*)", "\\1", x))
condition %<>% factor
rs <- rowSums(cts)
cts <- cts[rs > 100, ]
logcts <- log10(cts + 1)
```

After running the setup, your environment contains:

| Object      | Type                    | Description                                              |
|-------------|-------------------------|----------------------------------------------------------|
| `se`        | SummarizedExperiment    | Scaled GTEx brain RNA-seq; all genes, all samples        |
| `cts`       | integer matrix          | Count matrix filtered to genes with rowSum > 100         |
| `logcts`    | numeric matrix          | log10(cts + 1)                                           |
| `condition` | factor                  | Brain region label for each sample                       |

Check the dimensions of each object:

```{r}
dim(se)
dim(cts)
dim(logcts)
length(condition)
nlevels(condition)
table(condition)
```

---

## Exercise 1: Distributions and region comparison

**1a.** How many unique brain regions are in this dataset, and how many
samples belong to each? Use `levels()` and `table()` to explore
`condition`.

```{r}

```

**1b.** Pick **three** brain regions that were *not* analyzed in
`brain_RNA.qmd` (which used Cerebellum, Cortex, Hippocampus, and
Hypothalamus). Subset `logcts` and `condition` to only those regions,
dropping unused factor levels.

```{r}

```

**1c.** For each sample in your subset, compute the median log-count
across all genes. Build a `data.frame` with columns `region` and
`med`, then use ggplot2 to make a boxplot or violin plot of per-sample
median log-count by region.

*Hint: `apply(mat, 2, median)` gives column medians.*

```{r}

```

**1d.** Find the single gene with the highest row variance **across
all ~1400 samples** (use `rowVars` from `matrixStats`). Plot its
log-count values across your three regions from 1b as a ggplot2
boxplot. Does its expression differ by region?

```{r}

```

**1e.** Pick two of your three regions and compute the mean
log-expression of each gene within each region (hint: `rowMeans` on
the appropriate column subset). Make a `smoothScatter` of region A
means vs. region B means. Add a diagonal reference line with
`abline(0, 1)`. What does the spread around the diagonal tell you?

```{r}

```

---

## Exercise 2: PCA on a new subset of regions

**2a.** Using the same three regions from Exercise 1b, identify the top
500 genes by row variance computed **only across those samples**. Run
`prcomp` on the transposed log-count matrix for those genes and
samples.

```{r}

```

**2b.** Make a ggplot2 scatter plot of PC1 vs. PC2, coloring points by
brain region. How much variance does each PC explain? (Check
`summary(pc)` or compute from `pc$sdev`.) How does the separation
compare to what you saw in `brain_RNA.qmd` with the original four
regions?

```{r}

```

**2c.** What are the top 10 most variable genes (by `rowVars`) among
your subset from 2a? Print their Ensembl IDs — these are the row names
of `logcts`. Look one up on [Ensembl](https://ensembl.org) or
[NCBI Gene](https://www.ncbi.nlm.nih.gov/gene/). Does it make
biological sense that this gene varies across brain regions?

```{r}

```

**2d.** *(Extension)* How sensitive is the PCA structure to your choice
of top genes? Re-run the PCA using the top 100 and top 2000 genes by
variance and compare the PC1 vs. PC2 plots. Does the grouping by
region change?

```{r}

```
