---
title: "Simple EDA using dplyr and ggplot2"
author: "[Michael Love](http://mikelove.github.io)"
format:
  html:
    embed-resources: true
---

In this document, we will explore a small table of data about
the genomes of various organisms. I also include a file, 
`scrape_ensembl.R` which shows how I scraped this data
from the [Ensembl](http://ensembl.org) website.

We start by reading in the TSV (tab-separated value) data
using the readr package:

```{r message=FALSE}
library(readr)
library(here)
orgs <- read_tsv(here("eda","organisms_genes.tsv"))
head(orgs)
table(orgs$type)
```

`read_tsv` returns a *tibble* object, which is similar to a
*data.frame*, being tabular data with different types of variables
as columns (character, double, integer, etc.). One difference, aside
from the presentation when you `show` the object, is that the
characters are not automatically converted to *factor*, as they are
with `read.delim` or the like.

Let's convert the first column to a *factor*:

```{r message=FALSE}
orgs$type <- factor(orgs$type)
```

Let's make some plots of the number of basepairs in the genomes of
these organisms, split by the type of organism. There's one plant with
a really large genome. 
We could use the `stripplot` function, which is in the `lattice` package: 
this draws points for each level of a factor, similar to a boxplot,
but is preferred when there are not many observations (e.g. < 20).

```{r message=FALSE}
library(lattice)
with(orgs, stripplot(bp ~ type))
with(orgs, stripplot(log10(bp) ~ type))
```

Compare to boxplot. Again, given the small number of samples, `stripplot`
is preferred.

```{r}
with(orgs, boxplot(bp ~ type))
```

If you haven't seen `with(...)` before, it's a way to temporarily
use the variables (columns) of a *data.frame* as if they were
variables in the environment, that is, without using the `$`. As we
will see below, `ggplot2` also offers such a convenient syntax for
working with column names.

Most of the organisms have around 1 billion basepairs in their genome.

That large genome is wheat.

```{r}
orgs[which.max(orgs$bp),]
```

Let's remove it for now to make plots where we can see the range of
the other organisms.

```{r}
orgs <- orgs[orgs$name != "Wheat",]
```

I plot the number of genes that code for proteins over the total
number of basepairs:

```{r message=FALSE}
with(orgs, plot(bp, coding, col=type, pch=20, cex=2))
legend("topright", levels(orgs$type), col=1:nlevels(orgs$type), pch=20)
# right click to Esc from identify()
# with(orgs, identify(bp, coding, labels=orgs$name)) 
```

Another way to make a plot with labels is using
[ggplot2](http://ggplot2.tidyverse.org/reference/).
In ggplot2, you build a plot by adding layers with `+`: a call to
`ggplot()` sets up the coordinate system and maps columns to visual
properties via `aes()`, and then `geom_*` functions draw the data.
Common geoms include `geom_point()`, `geom_line()`, `geom_histogram()`,
`geom_density()`, `geom_boxplot()`, and `geom_smooth()`.

```{r}
library(ggplot2)
ggplot(orgs, aes(bp, coding, color=type)) +
  geom_point(size=3)
```

Which organisms have the largest genome (after wheat)?

```{r}
orgs[order(orgs$bp, decreasing=TRUE)[1:5],c("name","bp")]
```

Another way to get quick summaries of a dataset is `skim` from the 
*skimr* package:

```{r}
library(skimr)
skim(orgs)
```

For many operations, it's useful to use the dplyr package to
manipulate *data.frames* or *tibbles*. 

Before I show dplyr operations, let me introduce the pipe operator `|>`,
which sends the left-hand side to the first argument of the function on
the right-hand side.

A simple example of the pipe:

```{r}
x <- c(1:9,1000)
x |> mean(trim=.1)
# equivalent to
mean(x, trim=.1)
```

Using dplyr, let's add a new column tallying up coding and non-coding genes:

```{r message=FALSE}
library(dplyr)
orgs <- orgs |> mutate(genes = coding + noncoding)
orgs$genes
```

dplyr is really useful for grouping and summarization tasks. The
following line groups the organisms by type and summarizes the number
for each type, the mean ratio of genes which are coding and the
standard deviation of the ratio. One `type` has an SD of NA, why?

```{r}
orgs |>
  group_by(type) |>
  summarize(n=n(),
            cr_avg=mean(coding/genes),
            cr_sd=sd(coding/genes))
```

Let's save this as a new table, `tab`:

```{r}
tab <- orgs |>
  group_by(type) |>
  summarize(n=n(),
            cr_avg=mean(coding/genes),
            cr_sd=sd(coding/genes))
```

Two packages with functions for printing tables (as Latex or
markdown):

```{r}
library(xtable)
xtable(tab)
library(knitr)
kable(tab)
```

We can use summarization as a mid-point and then send those summaries
to a plotting function. For example, points and line ranges:

```{r}
# here we have to name the median something other than `bp`
orgs |>
  group_by(type) |>
  summarize(basepairs=median(bp),
            min=min(bp),
            max=max(bp)) |>
  ggplot(aes(type, basepairs, ymin=min, ymax=max)) +
  geom_pointrange()
```

ggplot2 is great for grouping points together with summarizing
graphical elements, such as lines through sets of points:

```{r}
ggplot(orgs, aes(bp, coding, col=type)) +
  geom_point() + 
  geom_smooth(se=FALSE, method="lm", show.legend=FALSE)
```

Also for "faceting" plots, breaking into multiple scatter plots by
breaking along a variable:

```{r}
ggplot(orgs, aes(bp, coding)) +
  geom_point() + 
  facet_wrap(~ type)
```

Finally, I like to show a useful plot constructed using ggplot2, just
with simulated data. It is common that wants to show overlapping
densities or histograms, which is not a simple task with base R
graphics. It's fairly simple with ggplot2.

We create some simulated data drawing three sets of samples from three
distributions. 

```{r}
ns <- c(1000,500,200)
dat <- data.frame(type=rep(letters[1:3], ns),
                  z=rnorm(1700,
                          mean=rep(c(0,2,4),ns),
                          sd=rep(c(1,1,1),ns)))
head(dat)
table(dat$type)
```

It's straightforward to draw these as overlapping densities:

```{r}
ggplot(dat, aes(z, col=type, fill=type)) + geom_density(alpha=0.1)
```

Or as histograms:

```{r}
ggplot(dat, aes(z, col=type, fill=type)) +
  geom_histogram(alpha=0.1, position="identity")
```

Another option, putting the bars beside each other:

```{r}
ggplot(dat, aes(z, col=type, fill=type)) +
  geom_histogram(position="dodge")
```

## Appendix: `|>` vs `%>%` and the placeholder

R 4.1 introduced a native pipe `|>` that covers the most common use
case: passing the left-hand side to the *first* argument of the
right-hand side. For almost everything in this course, `|>` and `%>%`
(from the *magrittr* package) are interchangeable.

The one place they differ is the placeholder, a way to pipe into
an argument that is not the first, or that is nested inside another
call.

Base R uses `_` as a placeholder, but only as a direct named argument.

```{r message=FALSE}
library(magrittr)
orgs |> lm(coding ~ bp, data = _) |> coef() # base R
orgs %>% lm(coding ~ bp, data = .) %>% coef() # magrittr equivalent
orgs %>% lm(coding ~ bp, data = .) %>% coef # magrittr can drop ()
```

Magrittr's `.` can appear anywhere on the right-hand side, including
nested inside another call. Here we compute `trim` from the data itself:

```{r}
x <- c(1:9, 1000)
x |> mean(trim = 0.1)
x %>% mean(trim = 1/length(.))  # trim = 1/10 = 0.1
# x |> mean(trim = 1/length(_)) # syntax error
```


Magrittr also provides `%<>%`, a compound assignment pipe that applies
a function and reassigns the result to the left-hand side. There is no
base R equivalent; the plain `<-` form is always a safe replacement:

```{r}
# these two are equivalent
orgs$type %<>% factor       # magrittr compound assignment
orgs$type <- factor(orgs$type)  # base R
```
