---
title: "Quick primer on NA, Inf, and NaN"
author: "[Michael Love](http://mikelove.github.io)"
output: html_document
---

When performing data analysis with large datasets or public data, it's
important to be aware of the types of missing values that might be
present in the data, and might accumulate as you compute summary
statistics. 

Many already will know about the `NA` value, which propagates into
summary statistics like so:

```{r}
x <- c(1,2,3,NA)
mean(x)
sd(x)
```

And can be removed from those statistics either through the function
or manually using `is.na`:

```{r}
mean(x, na.rm=TRUE)
sd(x, na.rm=TRUE)
is.na(x)
mean(x[!is.na(x)])
```

When we have constant values in a matrix, and we compute z-scores, we
can get two kinds of output, either 0/0 which is "not a number" `NaN`, 
or any non-zero number divided by 0 which gives `Inf` or `-Inf`:

```{r}
m <- matrix(c(0,0,0,1,1,1,1:9),ncol=3,byrow=TRUE)
m
rowMeans(m)
apply(m, 1, sd)
z <- apply(m, 1, function(x) mean(x)/sd(x))
z
```

We can identify the `NaN` and infinite values using `is.nan` and
`is.finite`: 

```{r}
z[!is.nan(z)]
z[!is.nan(z) & is.finite(z)]
```

We could also deal with these by explicitly checking if the
denominator is equal to 0 and returning whatever we want in this case: 

```{r}
apply(m, 1, function(x) if (sd(x) == 0) 0 else mean(x)/sd(x))
```
