---
title: "Homework 0 - Basic inference and GitHub submission"
author: "your name here"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: html_document
---

# Question 1

Given $X_i$ where $i = 1, ..., 10$, a random sample from a
distribution with cdf 

$$ F(x) = 1 - e^{-a x} $$

for $x \ge 0$ and zero otherwise, derive and calculate for the given
sample the maximum likelihood estimate (MLE) of $a$. 

```{r}
set.seed(1)
a <- 5
x <- rexp(10, rate=a)
```

$$ f(x) = \dots $$

$$ L(a | X_1, ..., X_{10}) = \dots $$

# Question 2

Suppose we have a bivariate distribution with 

$$f(x, y) = x y^2 / 13 $$

with support $\{(1,1), (1,2), (2,2)\}$. Compute $E(X|Y=2)$

```{r echo=FALSE}
f <- function(x,y) x * y^2 / 13
feval <- c(f(1,1), f(1,2), f(2,2))
barplot(feval, names.arg=c("(1,1)","(1,2)","(2,2)"),
        ylim=c(0,1), xlab="(x,y)", ylab="density")
```

# Question 3

There are three coins in a bag, two are fair with $P(\textrm{heads}) =
0.5$ and one is biased with $P(\textrm{heads}) = 0.75$. Suppose we
pull a coin out and flip it 4 times, getting heads 3 times. What is
the posterior probability the coin was the biased one?

```{r}
coin <- c(1,1,0,1)
```

$~$

$~$

$~$
